---
title: "Plastic Degredation Term Project"
author:
- Heather MacTavish
- Andrew Forrest
- Yvanna Tchatchoua
- Laurel Genge
- Carlie Barnhill
- Max Berthold
- Mireille Savoie
- Douglas A. Campbell
date: "`r format(Sys.Date())`"
output:
  bookdown::html_document2:
    code_folding: show
    keep_md: yes
    toc: TRUE
    toc_float: TRUE
    toc_depth: 6
    fig_caption: yes
bibliography: ../Docs/reference_plastic.bib
csl:  ../Docs/apa-with-abstract.csl
editor_options: 
  markdown: 
    wrap: 72
---

XXX what should the knit output be? XXX


# Acknowledgement
This R markdown file is adapted from the "Import_MCData" file created by Laurel Genge, Carlie Barnhill, Max Berthold, Mireille Savoie, and Douglas A. Campbell for the Plastic Degredation project by Heather MacTavish, Andrew Forrest, and Yvanna Tchatchoua in BIOL3111.

Dr. Campbell was supported by the Canada Research Chairs.

# Set Options {-}
## Set figure caption font size {-}
```{css, echo=FALSE}
p.caption {
  font-size: 18px;
}
```
## Set Chunk Options {-}
```{r setup2, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE)
knitr::opts_chunk$set(fig.path='Figs/')
```


XXX get Yvanna and andrew to add their parts to this file






XXX write that this was by the original authors
# Introduction 

The PSI Multicultivator is used to grow 8 x 80 ml of phytoplankton
culture under a common temperature regime, with individual control of
bubbling, light level, light spectral quality and photoperiod for each
of the 8 culture tubes.

This .Rmd Rworkbook imports data in simple .csv long form exported from
PSI Multicultivators based upon project specific values for variables
set by the user.

It tidies and organizes the data. It uses a pivot_wider and
interpolation approach to get the Actinic_par and OD values in line
rowwise. This requires careful 'arrange' of the rows. It imports a
metadata catalog (Data/RawData/"CultureCatalog.Rds") and merges the metadata with the imported data based
upon shared values for the variables 'MC', 'Tube', and 'Filename' which
should unambiguously identify a given growth trajectory measured at
OD680 or OD720. OD680 of 0.005 and OD720 of 0.003 were used to replace measured OD values for tubes with OD680 and/or OD720 < 0 across the whole growth trajectory. Assigning these very low OD values acts as a place holder; otherwise data for these tubes get filtered out later and 'no growth' data are lost.

It generates preliminary data plots. It filters the data for outliers by
screening out values distant from the moving average of a window in the
stream; this requires careful 'arrange' of the rows so sequential rows
represent sequential time steps.

This works because the OD680 & OD720 data are only episodically, but
widely, aberrant when a bubble interrupts the measurement, and if the
Multicultivator is running properly these bubble aberration events are
rare.

# Load Libraries
```{r load libraries, echo = FALSE, warning = FALSE, message = FALSE}
library(tidyverse)
library(lubridate)
library(broom)
library(data.table)
library(zoo)
#library(lme4) #XXX
#library(multcomp) #XXX
```

# Set Project Variables
```{r set project variables}
Project <- "Plastic"


# Multicultivator data files are too large to upload in a single .zip therefore 3 .zipped folders are uploaded. 

DataIn <- file.path("..", "Data", "RawData", "MultiCultiPlasticData.zip")

DataOut <- file.path("..","Data", "ProcessedData", "ImportedMCData")

# number of rows to skip upon read csv
Skip <- 20
```

#set Photobiology colors XXX
```{r set colours}
# Set colour codes for each WL of the Multicultivator. These colours are matched to WL using the Photobiology package. Colour codes used for visualization of spectral settings used in Muticulti run.

MyWavelengths = c(405, 450, 470, 530, 620, 660, 730, "WW")
MCMIXColours = c(c("purple", "blue", "skyblue", "green2",           "yellow", "red", "darkred", "black"))


names(MCMIXColours) <- MyWavelengths
MCMIXColours

```

```{r read locally stored metadata from rds}

CultureCatalog <- readRDS(file = file.path("..", "Data",  "RawData","CultureCatalog.Rds"))

```


# List previously imported MultiCulti files
```{r previously imported files}
list.files(path = DataOut, pattern = Project, full.names = TRUE)
```
# BioReactor Data

### List MultiCulti files for Project that are saved in DataIn
```{r MultiCulti files}

#List files in the extracted folder with a ".asc" extension
MCFiles <- unzip(DataIn, list = TRUE)
MCFiles <- MCFiles[grepl(".csv$", MCFiles$Name), "Name"]
print(MCFiles)

```
## MCMIX004
### Set Target File and read in csv
```{r set TargetFile and read csv MCMIX004}

TargetFile <- "20240927_Plastic_MCMIX004.csv"

# This opens a compressed file without decompressing.
TargetData <- read_csv(unz(DataIn, TargetFile),  
                       skip = Skip, 
                       id = "Path", 
                       col_names = c("key", "time", 
                                     "abs-time", "value")) %>% 
  separate(col =Path, into = c("Path", "Filename"), sep = ":") 

TargetFileName <- str_remove(string = TargetFile, pattern = ".csv")
TargetFileName
```

### Filter superfluous rows, Add ToD column, extract ExpDate, extract MC, 
```{r tidy MultiDataTarget MCMIX004}
#filter superfluous rows to simplify later pivot
TargetData <- TargetData %>%
  filter(str_detect(key, "od-720|od-680|actinic-lights.light")) %>%
  dplyr::select(key, time, `abs-time`, value, Filename) %>% 
  mutate(abs_time = mdy_hms(`abs-time`),
    Tube = str_extract(key, "-[:digit:]"),
         Tube = as.numeric(str_extract(Tube, "[:digit:]"))
         ) %>%
  dplyr::select(-`abs-time`)


#extract StartHour dynamically from first row of abs_time and display for cross check
StartHour <- as.numeric(format(TargetData$abs_time[1], format = "%H"))
StartHour
 
StartDate <- TargetData$abs_time[1]
StartDate

#Generate ToD as mod 24 of time + StartHour
TargetData <- TargetData %>%
  mutate(ToD = (time + StartHour) %% 24,
         Day = round((time/24), digits = 0))
#had trouble extracting 'Day' from abs_time, only got 3 unique values
  
#extract ExpDate for matching with Catalog
TargetData <- TargetData %>% 
    mutate(ExpDate = str_extract(Filename, "202[:digit:][:digit:][:digit:][:digit:][:digit:]_"),
           ExpDate = ymd(str_extract(ExpDate, "202[:digit:][:digit:][:digit:][:digit:][:digit:]")))

#extract MC for matching with Catalog
#fixed to run with MC or MCMIX.
#fixed to work for RUN #'s > 100. MC extracted from filename when Runs >99 returns the MC name and RUN #, this removes '_RUN#'
TargetData <- TargetData %>% 
    mutate(MC = str_extract(Filename, "MC.*[:digit:][:digit:][:digit:]"),
           MC = str_remove(MC, pattern = "\\_.*"))
```

## Create preliminary plot for TargetData (MCMIX004)
### Plots all OD values. True detection is OD680 and false detection is OD720.
```{r prelim plot MCMIX004}

  TargetData %>%
  filter(grepl("od-", key)) %>%
  ggplot(data = .) +
  geom_point(aes(x = time, y = value, colour = as.factor(str_detect(key, "680"))), size = 0.5) +
  geom_vline(aes(xintercept = 7 * 24), linetype = "dashed") + 
  geom_vline(aes(xintercept = 14 * 24), linetype = "dashed") + 
  scale_colour_manual(values = c("black", "red")) +
  labs(y = "Optical Density (OD)", x = "Elapsed Time (h)", title = "Tubes") +
  facet_grid(cols = vars(as.factor(Tube))) +
  theme_bw()

 TargetData %>%
  filter(grepl("od-", key)) %>%
  ggplot(data = .) +
  geom_point(aes(x = time, y = value, colour = as.factor(str_detect(key, "680"))), size = 0.5) +
   coord_cartesian(ylim = c(-0.1,0.1)) +
  geom_vline(aes(xintercept = 7 * 24), linetype = "dashed") + 
  geom_vline(aes(xintercept = 14 * 24), linetype = "dashed") + 
  scale_colour_manual(values = c("black", "red")) +
  labs(y = "Optical Density (OD)", x = "Elapsed Time (h)", title = "Tubes") +
  facet_grid(cols = vars(as.factor(Tube))) +
  theme_bw() 


```

### Generate par_ue column with rows aligned with OD measures
Pivot_wider to get actinic-lights data aligned with relevant sensor
data. Need to include arrange(Filename, time, Tube) to keep things aligned! Need to
group_by and/or reorder rows appropriately; Be Careful

```{r pivot_wider MCMIX004}
#possible issue with data structure; there are multiple values for some of the rows of actinic light columns, so the column becomes a list.
#Can add  values_fn = 
#to arbitrarily take the max or min etc. element of the list; but there might be a wider problem here when importing multiple files

TargetDataWide <- TargetData %>%
  pivot_wider(names_from = key, values_from = value, values_fn = list(value = max)) %>%
  arrange(Filename, MC, Tube, time)

TargetDataWide[1:10,]

rm(TargetData)
```

Actinic light values do not align time wise with OD measures.
Interpolate NA in actinic light columns from last observation, arrange
by MC & Tube Then generate Actinic_par summary column. If multiple lights
are activated, this chunk will give the summed par of all different
colours for the tube. If a single actinic light is activated per tube,
this gives the par for that tube. Filter rows where !is.na(Actinic_par)
to check for incorrect row sums.

Interpolation for Sine is not necessarily appropriate interpolation for
Square photoregime; issues with propagating last Actinic_par of
afternoon through evening, or back-casting first Actinic_par of morning.

Small glitching adding actinic_light values for tubes where
actinic_light column should be 0; issue with interpolation we think.

```{r interpolate and summarize actinic_par by tube MCMIX004}
#http://publish.illinois.edu/spencer-guerrero/2014/12/11/2-dealing-with-missing-data-in-r-omit-approx-or-spline-part-1/
#https://dplyr.tidyverse.org/dev/articles/colwise.html
#Interpolation causes problems with final rows that repeat last value.

interpolate <- function(x){zoo::na.locf(x, na.rm = FALSE, fromLast = FALSE, type = "l", maxgap = Inf)}

#possible problem with actinic_par for MC data b/c actinic-lights.light1 = NA generates actinic_par of 0 b/c in rowSums na.rm = TRUE, which treats NA as 0.
#possibly not a big problem but watch for bugs
#na.rm = FALSE fails to run
TargetDataWide <- TargetDataWide %>%
  group_by(Tube) %>%
  arrange(Filename, MC, Tube, time) %>%
  mutate(across(.cols = starts_with("actinic-lights.light"), .fns = interpolate)) %>%
  ungroup() %>%
  mutate(Actinic_par = rowSums(.[grep("actinic-lights.light", names(.))], na.rm = TRUE)) %>%
  filter(!is.na(Actinic_par)) %>%
   dplyr::select(!contains("actinic-lights.light"))


TargetDataWide[1:10,]
```

Now that Actinic_par is aligned with each row, coalesce
od-sensors-X.od-720 and od-sensors-X.od-680 into 2 columns, b/c 'Tube'
is already a column, so no need to identify tube X in
od-sensors-X.od-680 columns. This might cause problems later matching OD
measures to actinic light colours.

```{r consolidate OD MCMIX004}
TargetDataWide <- TargetDataWide  %>%
   mutate(OD680 = rowSums(.[grep("od-680", names(.))], na.rm = TRUE),
          OD720 = rowSums(.[grep("od-720", names(.))], na.rm = TRUE)) %>%
   dplyr::select(!contains("od-sensors"))

TargetDataWide[1:10,]

```

## Merge Data with meta data
```{r metadata merge MCMIX004}
#This generates 'NA' values for ~1,000,000 rows of 'O2'; possibly the temperature rows?
TargetDataMeta <- left_join(x = TargetDataWide, y= CultureCatalog, by = c("MC", "Tube"))

TargetDataMeta

rm(TargetDataWide)
```
## Prelim plot with culture meta (MCMIX004)
```{r second prelim plot MCMIX004}

  TargetDataMeta %>% ggplot() +
  geom_point(aes(x = time, y = OD680, colour = as.factor(WL)), size = 0.15) +
  geom_point(aes(x = time, y = OD720), size = 0.15, colour = "orange") +
  #geom_point(aes(x = time, y = Actinic_par/1000),  colour = "orange", size = 0.0001) +
  geom_vline(aes(xintercept = 7 * 24), linetype = "dashed") + 
  geom_vline(aes(xintercept = 14 * 24), linetype = "dashed") + 
  #scale_x_continuous(breaks=seq(0, 800, by = 125)) +
  #coord_cartesian(xlim = c(-10, 800)) +
 # scale_colour_manual(values = MCMIXColours) +
  #labs(y = "Optical Density (OD)", x = "Elapsed Time (h)", subtitle = "Growth Light (µE); Strain; SampleID; Tube") +
  facet_grid(rows = vars(as.factor(Plastic_Present)), cols = vars(as.factor(Tube), Nutrient_Present, Culture_Present)) +
  theme_bw()

```


## MCMIX006
### Set Target File and read in csv
```{r set TargetFile and read csv MCMIX006}

TargetFile <- "20240927_Plastic_MCMIX006.csv"


# This opens a compressed file without decompressing.
TargetData <- read_csv(unz(DataIn, TargetFile),  skip = Skip, id = "Path", col_names = c("key", "time", "abs-time", "value")) %>% 
  separate(col =Path, into = c("Path", "Filename"), sep = ":") 



TargetFileName <- str_remove(string = TargetFile, pattern = ".csv")

TargetFileName

```

### Filter superfluous rows, Add ToD column, extract ExpDate, extract MC,
```{r tidy MultiDataTarget MCMIX006}
#filter superfluous rows to simplify later pivot
TargetData <- TargetData %>%
  filter(str_detect(key, "od-720|od-680|actinic-lights.light"))
#head(TargetData)

# some files have "V5"  instead of "value" column
TargetData <- TargetData %>%
  dplyr::select(key, time, `abs-time`, value, Filename) %>% #, CDateTime
  mutate(Tube = str_extract(key, "-[:digit:]"),
         Tube = as.numeric(str_extract(Tube, "[:digit:]")),
         abs_time = mdy_hms(`abs-time`)) %>%
  dplyr::select(-`abs-time`)


#extract StartHour dynamically from first row of abs_time and display for cross check
StartHour <- as.numeric(format(TargetData$abs_time[1], format = "%H"))
StartHour
 
StartDate <- TargetData$abs_time[1]
StartDate

#Generate ToD as mod 24 of time + StartHour
TargetData <- TargetData %>%
  mutate(ToD = (time + StartHour) %% 24,
         Day = round((time/24), digits = 0))
#had trouble extracting 'Day' from abs_time, only got 3 unique values
  
#extract ExpDate for matching with Catalog
TargetData <- TargetData %>% 
    mutate(ExpDate = str_extract(Filename, "202[:digit:][:digit:][:digit:][:digit:][:digit:]_"),
           ExpDate = ymd(str_extract(ExpDate, "202[:digit:][:digit:][:digit:][:digit:][:digit:]")))

#extract MC for matching with Catalog
#fixed to run with MC or MCMIX.
#fixed to work for RUN #'s > 100. MC extracted from filename when Runs >99 returns the MC name and RUN #, this removes '_RUN#'
TargetData <- TargetData %>% 
    mutate(MC = str_extract(Filename, "MC.*[:digit:][:digit:][:digit:]"),
           MC = str_remove(MC, pattern = "\\_.*"))

```

## Create preliminary plot for TargetData MCMIX006
### Plots all OD values. True detection is OD680 and false detection is OD720.
```{r prelim plot MCMIX006}

  TargetData %>%
  filter(grepl("od-", key)) %>%
  ggplot(data = .) +
  geom_point(aes(x = time, y = value, colour = as.factor(str_detect(key, "680"))), size = 0.5) +
  geom_vline(aes(xintercept = 7 * 24), linetype = "dashed") + 
  geom_vline(aes(xintercept = 14 * 24), linetype = "dashed") + 
  scale_colour_manual(values = c("black", "red")) +
  labs(y = "Optical Density (OD)", x = "Elapsed Time (h)", title = "Tubes") +
  facet_grid(cols = vars(as.factor(Tube))) +
  theme_bw()

 TargetData %>%
  filter(grepl("od-", key)) %>%
  ggplot(data = .) +
  geom_point(aes(x = time, y = value, colour = as.factor(str_detect(key, "680"))), size = 0.5) +
   coord_cartesian(ylim = c(-0.1,0.1)) +
  geom_vline(aes(xintercept = 7 * 24), linetype = "dashed") + 
  geom_vline(aes(xintercept = 14 * 24), linetype = "dashed") + 
  scale_colour_manual(values = c("black", "red")) +
  labs(y = "Optical Density (OD)", x = "Elapsed Time (h)", title = "Tubes") +
  facet_grid(cols = vars(as.factor(Tube))) +
  theme_bw() 


```


### Generate par_ue column with rows aligned with OD measures

Pivot_wider to get actinic-lights data aligned with relevant sensor
data. Need to include arrange(Filename, time, Tube) to keep things aligned! Need to
group_by and/or reorder rows appropriately; Be Careful

```{r pivot_wider MCMIX006}
#possible issue with data structure; there are multiple values for some of the rows of actinic light columns, so the column becomes a list.
#Can add  values_fn = 
#to arbitrarily take the max or min etc. element of the list; but there might be a wider problem here when importing multiple files

TargetDataWide <- TargetData %>%
  pivot_wider(names_from = key, values_from = value, values_fn = list(value = max)) %>%
  arrange(Filename, MC, Tube, time)

TargetDataWide[1:10,]

rm(TargetData)
```

```{r interpolate and summarize actinic_par by tube MCMIX006}
#http://publish.illinois.edu/spencer-guerrero/2014/12/11/2-dealing-with-missing-data-in-r-omit-approx-or-spline-part-1/

#https://dplyr.tidyverse.org/dev/articles/colwise.html

#Interpolation causes problems with final rows that repeat last value.

interpolate <- function(x){zoo::na.locf(x, na.rm = FALSE, fromLast = FALSE, type = "l", maxgap = Inf)}

#possible problem with actinic_par for MC data b/c actinic-lights.light1 = NA generates actinic_par of 0 b/c in rowSums na.rm = TRUE, which treats NA as 0.
#possibly not a big problem but watch for bugs
#na.rm = FALSE fails to run
TargetDataWide <- TargetDataWide %>%
  group_by(Tube) %>%
  arrange(Filename, MC, Tube, time) %>%
  mutate(across(.cols = starts_with("actinic-lights.light"), .fns = interpolate)) %>%
  ungroup() %>%
  mutate(Actinic_par = rowSums(.[grep("actinic-lights.light", names(.))], na.rm = TRUE)) %>%
  filter(!is.na(Actinic_par)) %>%
   dplyr::select(!contains("actinic-lights.light"))


TargetDataWide[1:10,]
```

```{r consolidate OD MCMIX006}
TargetDataWide <- TargetDataWide  %>%
   mutate(OD680 = rowSums(.[grep("od-680", names(.))], na.rm = TRUE),
          OD720 = rowSums(.[grep("od-720", names(.))], na.rm = TRUE)) %>%
   dplyr::select(!contains("od-sensors"))

TargetDataWide[1:10,]

```

## Merge Data with meta data

```{r metadata merge MCMIX006}
#This generates 'NA' values for ~1,000,000 rows of 'O2'; possibly the temperature rows?
TargetDataMeta <- left_join(x = TargetDataWide, y= CultureCatalog, by = c("MC", "Tube"))

TargetDataMeta

rm(TargetDataWide)
```
# Prelim plot with culture meta data (MCMIX006)
```{r second prelim plot MCMIX006}

  TargetDataMeta %>% ggplot() +
  geom_point(aes(x = time, y = OD680, colour = as.factor(WL)), size = 0.15) +
  geom_point(aes(x = time, y = OD720), size = 0.15, colour = "orange") +
  #geom_point(aes(x = time, y = Actinic_par/1000),  colour = "orange", size = 0.0001) +
  geom_vline(aes(xintercept = 7 * 24), linetype = "dashed") + 
  geom_vline(aes(xintercept = 14 * 24), linetype = "dashed") + 
  #scale_x_continuous(breaks=seq(0, 800, by = 125)) +
  #coord_cartesian(xlim = c(-10, 800)) +
  #scale_colour_manual(values = MCMIXColours) +
  #labs(y = "Optical Density (OD)", x = "Elapsed Time (h)", subtitle = "Growth Light (µE); Strain; SampleID; Tube") +
  facet_grid(rows = vars(as.factor(Plastic_Present)), cols = vars(as.factor(Tube), Nutrient_Present, Culture_Present)) +
  theme_bw()

```






XXXLook at what you want for filtering
Fix meta catalog
do for both of the files
fix labels
as a formal practice segregate plastic (but fine for this project)
remove extra columns from meta





# Plastic Analyses
```{r plastic percent change}
# make a new data set to not mess up the original
plastic <- CultureCatalog


# Calculate percentage change
plastic$perc_change <- (plastic$Final_Plastic_Weight_mg - plastic$Initial_Plastic_Weight_mg) / plastic$Initial_Plastic_Weight_mg * 100

plastic
```
```{r plastic percent change graph}
ggplot(plastic, aes(x = as.factor(Culture_Present), y = perc_change, fill = as.factor(Nutrient_Present))) +
  geom_boxplot() +
  labs(title = "Plastic Degradation Across Conditions", x = "Culture Present", y = "Percentage Change (%)") +
  theme_minimal()
```


# Microbial Suspension and Wall Growth Analyses

```{r}
# Scatter plot: microbial growth vs plastic degradation
ggplot(plastic, aes(x = perc_change, y = Wall_Growth_RFU, color = as.factor(Nutrient_Present))) +
  geom_point() +
  labs(title = "Plastic Degradation vs Microbial Growth", x = "Percentage Plastic Change (%)", y = "Wall Growth RFU") +
  theme_minimal()
```

```{r}
# ANOVA model
anova_model <- aov(perc_change ~ Culture_Present * Nutrient_Present * Temperature_C, data = plastic)
summary(anova_model)

# plot of the interactions
interaction.plot(plastic$Temperature_C, plastic$Culture_Present, plastic$perc_change,
                 col = 1:2, xlab = "Temperature", ylab = "Plastic Degradation (%)",
                 main = "Interaction Between Temperature and Culture")

```

XXX
```{r}
# Correlation matrix
library(reshape2)
cor_data <- cor(plastic[, c("perc_change", "Wall_Growth_RFU", "Suspension_RFU")], use = "complete.obs")
cor_melt <- melt(cor_data)

# Heatmap
ggplot(cor_melt, aes(Var1, Var2, fill = value)) +
  geom_tile() +
  scale_fill_gradient2(low = "blue", high = "red", mid = "white", midpoint = 0) +
  labs(title = "Correlation Heatmap", fill = "Correlation") +
  theme_minimal()

```




## converting the variables to factor type
```{r making variables factors}
# Convert grouping variables to factors 
plastic$Plastic_Present <- as.factor(plastic$Plastic_Present)
plastic$Nutrient_Present <- as.factor(plastic$Nutrient_Present)
plastic$Culture_Present <- as.factor(plastic$Culture_Present)
```




# Extra analyses
```{r percent loss or gain plastic}
graph_data <- CultureCatalog %>%
  mutate(Weight_Loss_Percent = ((Initial_Plastic_Weight_mg - Final_Plastic_Weight_mg) / Initial_Plastic_Weight_mg) * 100)

graph_data$Culture_Present <- as.factor(graph_data$Culture_Present)

head(graph_data)



```

# Yvanna
```{r}
data_long<-CultureCatalog%>%
  pivot_longer(cols=c(Initial_Plastic_Weight_mg,Final_Plastic_Weight_mg),names_to = "Weight_Type",values_to = "Weight")
data_long

#putting them in the right order
data_long$Weight_Type <- factor(data_long$Weight_Type, levels = c("Initial_Plastic_Weight_mg", "Final_Plastic_Weight_mg"))
```
```{r}
ggplot(data_long,
       aes(x=Tube,y=Weight,fill=Weight_Type))+
  geom_bar(stat="identity",position="dodge")+
  scale_fill_manual(
    values=c("Initial_Plastic_Weight_mg"="lightblue",
             "Final_Plastic_Weight_mg"="red2"))+
  labs(x="Tube",y="Weight(mg)",fill="Weight Type")+
  facet_wrap(~ Temperature_C, labeller = labeller(Temperature_C = label_both)) +
  theme_minimal() +
  theme(
    strip.background = element_rect(fill = "lightgray", color = "black"), 
    strip.text = element_text(size = 9, face = "bold"), 
    panel.border = element_rect(color = "black", fill = NA, linewidth = 0.8)) 
  
```



```{r}


ggplot(graph_data, aes(x = Temperature_C, y = Wall_Growth_RFU, colour = Culture_Present)) +
  geom_point(size = 3, alpha = 0.7) +
  labs(x = "Temperature (°C)", y = "Wall Growth RFU", 
       title = "Scatter Plot of Wall Growth RFU vs Temperature") +
  theme_minimal()
ggplot(graph_data, aes(x = Temperature_C, y = Suspension_RFU, colour = Culture_Present)) +
  geom_point(size = 3, alpha = 0.7) +
  labs(x = "Temperature (°C)", y = "Suspension RFU", 
       title = "Scatter Plot of Suspension RFU vs Temperature") +
  theme_minimal()



```



# Andrew
I will be running an ANOVA (one way) on change in plastic mass data (initial - end) for experimental groups
(5 tubes each) in 15 and 25 degrees celsius.

This will tell us whether the mean '% change in plastic weight' is significantly different between the 15 and 25 degree groups


```{r}
filtered_dataplasticpresent<-CultureCatalog[CultureCatalog$Plastic_Present != 0,]
```


```{r}
onewayanova_mass<-aov((filtered_dataplasticpresent$Initial_Plastic_Weight_mg-filtered_dataplasticpresent$Final_Plastic_Weight_mg)~filtered_dataplasticpresent$Temperature_C)
summary(onewayanova_mass)
```

Factorial ANOVA using temperature, culture presence, and nutrient presence as independent variables.
This will tell us whether any one of these variables or some combination have a statistically significant effect on change in plastic mass.
```{r}
factorialanova_mass<-aov((filtered_dataplasticpresent$Initial_Plastic_Weight_mg-filtered_dataplasticpresent$Final_Plastic_Weight_mg)~filtered_dataplasticpresent$Temperature_C*filtered_dataplasticpresent$Culture_Present*filtered_dataplasticpresent$Nutrient_Present)
summary(factorialanova_mass)
```

One way ANOVAs for chlorophyll assay readings using temp as independent variable (one ANOVA for wall growth and another for suspension)...
```{r}
onewayanova_rfuwall<-aov(CultureCatalog$Wall_Growth_RFU~CultureCatalog$Temperature_C)
summary(onewayanova_rfuwall)
onewayanova_rfususpension<-aov(CultureCatalog$Suspension_RFU~CultureCatalog$Temperature_C)
summary(onewayanova_rfususpension)
```
Factorial ANOVAs for chlorophyll assay readings using temp, nutrient presence, plastic presence and culture presence as independent variables...
```{r}
factorialanova_rfuwall<-aov(CultureCatalog$Wall_Growth_RFU~CultureCatalog$Temperature_C*CultureCatalog$Culture_Present*CultureCatalog$Nutrient_Present*CultureCatalog$Plastic_Present)
summary(factorialanova_rfuwall)
factorialanova_rfususpension<-aov(CultureCatalog$Suspension_RFU~CultureCatalog$Temperature_C*CultureCatalog$Culture_Present*CultureCatalog$Nutrient_Present*CultureCatalog$Plastic_Present)
summary(factorialanova_rfususpension)
```










# more analyses 
XXX fix these
```{r}
# Summary statistics for Final_Plastic_Weight_mg
plastic %>%
  group_by(Culture_Present, Nutrient_Present, Plastic_Present, Temperature_C) %>%
  summarise(
    mean_wall_rfu = mean(Wall_Growth_RFU, na.rm = TRUE),
    mean_susp_rfu = mean(Suspension_RFU, na.rm = TRUE)
  )

```

```{r}
# Add a column for Percent Weight Loss
plastic <- plastic %>%
  mutate(
    Percent_Weight_Loss = ((Initial_Plastic_Weight_mg - Final_Plastic_Weight_mg) / Initial_Plastic_Weight_mg) * 100
  )
plastic <- plastic %>%
  mutate(
    Temperature_C = as.factor(Temperature_C),
    Culture_Present = as.factor(Culture_Present),
    Nutrient_Present = as.factor(Nutrient_Present)
  )


# Two-way ANOVA for Percent Weight Loss
weight_loss_anova <- aov(Percent_Weight_Loss ~ Temperature_C * Culture_Present * Nutrient_Present, data = plastic)
summary(weight_loss_anova)

# Tukey HSD for pairwise comparisons
TukeyHSD(weight_loss_anova)

```

```{r}
# Two-way ANOVA for Wall_Growth_RFU
rfu_anova <- aov(Wall_Growth_RFU ~ Temperature_C * Culture_Present * Nutrient_Present, data = plastic)
summary(rfu_anova)

# Linear regression for RFU vs Percent Weight Loss
rfu_regression <- lm(Percent_Weight_Loss ~ Wall_Growth_RFU + Suspension_RFU, data = plastic)
summary(rfu_regression)

# Scatterplot of RFU vs Percent Weight Loss
ggplot(plastic, aes(x = Wall_Growth_RFU, y = Percent_Weight_Loss, color = Temperature_C)) +
  geom_point() +
  geom_smooth(method = "lm", se = TRUE) +
  labs(x = "Wall Growth RFU", y = "Percent Weight Loss")

```

```{r}
# T-test for weight loss by temperature
t.test(Percent_Weight_Loss ~ Temperature_C, data = plastic)

# Boxplot for Percent Weight Loss by Temperature
ggplot(plastic, aes(x = factor(Temperature_C), y = Percent_Weight_Loss)) +
  geom_boxplot() +
  labs(x = "Temperature (°C)", y = "Percent Weight Loss")

```




```{r}
# Calculate Summary Statistics
summary_stats <- plastic %>%
  group_by(Temperature_C, Culture_Present, Nutrient_Present) %>%
  summarise(
    mean_loss = mean(Percent_Weight_Loss, na.rm = TRUE),
    se_loss = sd(Percent_Weight_Loss, na.rm = TRUE) / sqrt(n()),
    .groups = "drop")

# Bar Plot of Average Percent Weight Loss
ggplot(summary_stats, aes(x = factor(Temperature_C), y = mean_loss, fill = Culture_Present)) +
  geom_bar(stat = "identity", position = position_dodge(0.9)) +
  geom_errorbar(aes(ymin = mean_loss - se_loss, ymax = mean_loss + se_loss),
                position = position_dodge(0.9), width = 0.2) +
  labs(x = "Temperature (°C)", y = "Mean Percent Weight Loss", fill = "Culture Present") +
  theme_minimal()

# mean loss having negative means that it actually grew

# Interaction Plot
interaction.plot(plastic$Temperature_C, plastic$Nutrient_Present, plastic$Percent_Weight_Loss)

```

```{r}
ggplot(summary_stats, aes(x = factor(Temperature_C), y = mean_loss, fill = Culture_Present)) +
  geom_bar(stat = "identity", position = position_dodge(0.9)) +
  geom_errorbar(aes(ymin = mean_loss - se_loss, ymax = mean_loss + se_loss),
                position = position_dodge(0.9), width = 0.2) +
  labs(x = "Temperature (°C)", y = "Mean Percent Weight Loss", fill = "Culture Present") +
  facet_wrap(~ Nutrient_Present) +
  theme_minimal()

```

```{r}
# Bar Plot with Facet Labels and Borders
ggplot(summary_stats, aes(x = factor(Temperature_C), y = mean_loss, fill = Culture_Present)) +
  geom_bar(stat = "identity", position = position_dodge(0.9)) +
  geom_errorbar(aes(ymin = mean_loss - se_loss, ymax = mean_loss + se_loss),
                position = position_dodge(0.9), width = 0.2) +
  labs(
    x = "Temperature (°C)",
    y = "Mean Percent Weight Loss",
    fill = "Culture Present"
  ) +
  facet_wrap(~ Nutrient_Present, labeller = labeller(Nutrient_Present = label_both)) +
  theme_minimal() +
  theme(
    strip.background = element_rect(fill = "lightgray", color = "black"), 
    strip.text = element_text(size = 12, face = "bold"), # Style facet labels
    panel.border = element_rect(color = "black", fill = NA, linewidth = 0.8) 
  )


```




### looking at the negative loss (gain) category
```{r negative loss (gain)}
# Filter the dataset for the negative mean category
negative_mean_data <- plastic %>%
  filter(Percent_Weight_Loss < 0)
negative_mean_data
```

XXX references


XXX need to do every single read me in every single folder





