---
title: "Plastic Degredation Term Project"
author:
- Heather MacTavish^1^*
- Andrew Forrest^1^*
- Yvanna Tchatchoua^1^*
- Douglas A. Campbell^1^*
- Mireille Savoie^1^*
- Laurel Genge^1^
- Carlie Barnhill^1^
- Max Berthold^1^

date: "`r format(Sys.Date())`"
output:
  bookdown::html_document2:
    code_folding: show
    keep_md: yes
    toc: TRUE
    toc_float: TRUE
    toc_depth: 6
    fig_caption: yes
bibliography: ../Docs/PBATDegradationCitations.bib
csl:  ../Docs/apa-with-abstract.csl
editor_options: 
  markdown: 
    wrap: 72
---
```{figure caption font size, echo=FALSE}
p.caption {
  font-size: 18px;
}
```

# Affiliations {-}
^1^Mount Allison University, New Brunswick, Canada  

*corresponding author 


XXX what should the knit output be? XXX
XXX need to update references + read mesXXX


# Acknowledgement {-}
This R markdown file is adapted from the `Import_MCData` file created by Laurel Genge, Carlie Barnhill, Max Berthold, Mireille Savoie, and Douglas A. Campbell [@genge_import_mcdata_2024] for the Plastic Degradation project by Heather MacTavish, Andrew Forrest, and Yvanna Tchatchoua in BIOL3111.

Dr. Campbell was supported by the Canada Research Chairs.


  

XXX get Yvanna and andrew to add their parts to this file

# Overview

According to Genge et al. (2024):

>This .Rmd Rworkbook imports data in simple .csv long form exported from PSI Multicultivators based upon project specific values for variables set by the user. 

>It tidies and organizes the data. It uses a pivot_wider and interpolation approach to get the Actinic_par and OD values in line rowwise. This requires careful 'arrange' of the rows. It imports a metadata catalog (Data/RawData/"CultureCatalog.Rds") and merges the metadata with the imported data based
upon shared values for the variables 'MC', 'Tube', and 'Filename' which should unambiguously identify a given growth trajectory measured at OD680 or OD720. OD680 of 0.005 and OD720 of 0.003 were used to replace measured OD values for tubes with OD680 and/or OD720 < 0 across the whole growth trajectory. Assigning these very low OD values acts as a place holder; otherwise data for these tubes get filtered out later and 'no growth' data are lost [@genge_import_mcdata_2024].

>It generates preliminary data plots. It filters the data for outliers by screening out values distant from the moving average of a window in the stream; this requires careful 'arrange' of the rows so sequential rows represent sequential time steps [@genge_import_mcdata_2024].

> This works because the OD680 & OD720 data are only episodically, but widely, aberrant when a bubble interrupts the measurement, and if the Multicultivator is running properly these bubble aberration events are rare [@genge_import_mcdata_2024].

This notebook is also used for data visualizations and statistical analyses related to the degradation of plastic and chlorophyll analyses.



XXX write that this was by the original authors
# Introduction 

According to Genge et al. (2024):

>The PSI Multicultivator is used to grow 8 x 80 ml of phytoplankton culture under a common temperature regime, with individual control of bubbling, light level, light spectral quality and photoperiod for each of the 8 culture tubes.



# Materials and Methods
## Sampling

At the Mount Allison University swan pond, two samples (1 liter each) of surface water from the pond’s edge at 4553’58’’ N, 6422’12’’ W were collected in sanitized plastic bottles. These samples were then transported back to the lab. 

## Preparation of Culture Nutrient

To satisfy minimal nutrient needs and to sustain the microbial community present in pond water samples, nitrate and phosphate solution was prepared. Ingredients included, 2mM disodium phosphate (144L) and 149.6g/L sodium nitrate (12.8mL). Final concentrations of 100M nitrate and 10M phosphate result in the final mixed solution of pond water and nutrient media. Each trial tube contained 80mL of pond water and 0.4mL of nutrient solution.

## Pre-trial Treatment and Weighing of Bioplastic Pieces

A Define Planet Veggie Based Compostable Poobag (SKU: 28946971) was cut into pieces of uniform composition and similar mass. Three holes were then poked through each fragment of plastic. Fragments were then washed in 70% ethanol for 5-10 seconds each to remove any extraneous material. Following ~30 minutes of drying, individual pieces were weighed in an aluminum boat on a Mettler Toledo MX5 microbalance to obtain pre-trial weights of each fragment. At this point fragments were numbered based on the Multi-Cultivator MC 1000-OD photobioreactor (Photon Systems Instruments) tube which they would end up in. Tubes and the plastic fragments they were to contain (if applicable) were labeled 1-16.

## Photobioreactor Trials

Prior to the addition of nutrient mix or pond water, bag fragments were fastened to glass oxygen delivery tubes within each of the two photobioreactors’ larger tubes. This was done by sliding the folded fragment up the oxygen tube via the holes poked beforehand. Next, 80mL of pond water and 0.4mL of nitrate-phosphate mix were dispensed into each photobioreactor tube (where applicable). Plastic fragments were adjusted to ensure that there was no immediate threat of fragments interfering with the sensors aligned with each tube. The glass housing in which the tubes were located was then filled with distilled water to the point where each reactor’s floating bob was about 1 inch from the low water sensor.

The settings used in photobioreactor 1 and photobioreactor 2 were identical apart from water temperature, which was 15C in photobioreactor 1 and 25C in photobioreactor 2 (meant to mimic water temperature in vivo in summer and fall), . Both bioreactors were set to 12 hours of white light (equal intensity)/12 hours of dark daily, equal oxygen bubbling, and to take measurements at 720nm (optical density) and 680nm (chlorophyll a & b) every 5 minutes. Once these settings had been input into the bioreactor’s software, the photobioreactor was turned on and trials began. 

Beginning the first day of incubation, the contents of each photobioreactor were suspended once weekly. During the trial period, multiple measurements were skewed due to interference of plastic or oxygen delivery tubes with sensors. Such interference issues were fixed as they were noticed by positioning the central oxygen delivery tube and/or the plastic away from sensors. 

## Post-trial Mass Measurement of Bioplastic Pieces

Following a 6-week trial period, the photobioreactors were shut down and plastic fragments were removed from their respective tubes and washed with water to remove any loosely connected mass. Fragments were then left to dry in open air for 24 hours. Post-trial masses of each fragment were taken with the Mettler Toledo MX5 microbalance and an aluminum boat. 

## Chlorophyll Assays

To measure wall growth and suspension growth of pond microbes during the trial period, non-acidified chlorophyll assays were performed on the contents of each tube. A total of 32 assays were performed: 16 for suspension growth and 16 for wall growth. Each wall growth assay used 5mL 90% 3:2 acetone: dimethyl sulfoxide solvent to detach microbes from tube walls by swirling. Approximately 2-3mL of solvent/microbe mix was transferred from each tube to its associated assay cuvette. Suspension growth pond water samples were pipetted into assay cuvettes following transfer to a new container and resuspension. Using glass pipettes, 2mL of solvent was distributed to each assay cuvette. Chlorophyll fluorescence readings were taken at excitation wavelength 43610nm and emission wavelength 68510nm using the Turner Trilogy fluorometer.  

YET TO BE WRITTEN: DATA HANDLING/ANALYSES

## Data Handling

## Set Chunk Options {-}
Formatted display of content from .md file on GitHub site.
Upon knitr figures will be saved to 'Figs/'
```{r setup, include = FALSE}
knitr::opts_chunk$set(warning = FALSE, error = FALSE, message = FALSE)
knitr::opts_chunk$set(fig.path='Figs/')
```

### Load Libraries
```{r load libraries, echo = FALSE, warning = FALSE, message = FALSE}
library(tidyverse)
library(lubridate)
library(broom)
library(data.table)
library(zoo)
#library(lme4) #XXX
#library(multcomp) #XXX
```

This report was generated in `R version 4.3.2` [@r_core_team_r_2023], with the packages: `tidyverse`[@wickham_welcome_2019], `lubridate` [@grolemund_dates_2011], `knitr` [@xie_knitr_2023; @xie_dynamic_2015; @stodden_knitr_2014], `broom` [@robinson_broom_2024],`data.table` [@barrett_datatable_2024], and `zoo` [@zeileis_zoo_2005].


### Set Project Variables
```{r set project variables}
Project <- "Plastic"

# Multicultivator data files are too large to upload in a single .zip therefore 3 .zipped folders are uploaded. 

DataIn <- file.path("..", "Data", "RawData", "MultiCultiPlasticData.zip")

DataOut <- file.path("..","Data", "ProcessedData", "ImportedMCData")

# number of rows to skip upon read csv
Skip <- 20
```


```{r read locally stored metadata from rds}
CultureCatalog <- readRDS(file = file.path("..", "Data",  "RawData","CultureCatalog.Rds"))
```

### List MultiCulti files for Project that are saved in DataIn
```{r MultiCulti files}

#List files in the extracted folder with a ".asc" extension
MCFiles <- unzip(DataIn, list = TRUE)
MCFiles <- MCFiles[grepl(".csv$", MCFiles$Name), "Name"]
print(MCFiles)

```
### MCMIX004 File
#### Set Target File and read in csv
```{r set TargetFile and read csv MCMIX004}

TargetFile <- "20240927_Plastic_MCMIX004.csv"

# This opens a compressed file without decompressing.
TargetData <- read_csv(unz(DataIn, TargetFile),  
                       skip = Skip, 
                       id = "Path", 
                       col_names = c("key", "time", 
                                     "abs-time", "value")) %>% 
  separate(col =Path, into = c("Path", "Filename"), sep = ":") 

TargetFileName <- str_remove(string = TargetFile, pattern = ".csv")
TargetFileName
```

#### Filter superfluous rows, Add ToD column, extract ExpDate, extract MC, 
```{r tidy MultiDataTarget MCMIX004}
#filter superfluous rows to simplify later pivot
TargetData <- TargetData %>%
  filter(str_detect(key, "od-720|od-680|actinic-lights.light")) %>%
  dplyr::select(key, time, `abs-time`, value, Filename) %>% 
  mutate(abs_time = mdy_hms(`abs-time`),
    Tube = str_extract(key, "-[:digit:]"),
         Tube = as.numeric(str_extract(Tube, "[:digit:]"))
         ) %>%
  dplyr::select(-`abs-time`)


#extract StartHour dynamically from first row of abs_time and display for cross check
StartHour <- as.numeric(format(TargetData$abs_time[1], format = "%H"))
StartHour
 
StartDate <- TargetData$abs_time[1]
StartDate

#Generate ToD as mod 24 of time + StartHour
TargetData <- TargetData %>%
  mutate(ToD = (time + StartHour) %% 24,
         Day = round((time/24), digits = 0))
#had trouble extracting 'Day' from abs_time, only got 3 unique values
  
#extract ExpDate for matching with Catalog
TargetData <- TargetData %>% 
    mutate(ExpDate = str_extract(Filename, "202[:digit:][:digit:][:digit:][:digit:][:digit:]_"),
           ExpDate = ymd(str_extract(ExpDate, "202[:digit:][:digit:][:digit:][:digit:][:digit:]")))

#extract MC for matching with Catalog
#fixed to run with MC or MCMIX.
#fixed to work for RUN #'s > 100. MC extracted from filename when Runs >99 returns the MC name and RUN #, this removes '_RUN#'
TargetData <- TargetData %>% 
    mutate(MC = str_extract(Filename, "MC.*[:digit:][:digit:][:digit:]"),
           MC = str_remove(MC, pattern = "\\_.*"))
```

#### Create preliminary plot for TargetData (MCMIX004)
##### Plots all OD values. True detection is OD680 and false detection is OD720.
```{r prelim plot MCMIX004}

  TargetData %>%
  filter(grepl("od-", key)) %>%
  ggplot(data = .) +
  geom_point(aes(x = time, y = value, colour = as.factor(str_detect(key, "680"))), size = 0.5) +
  geom_vline(aes(xintercept = 7 * 24), linetype = "dashed") + 
  geom_vline(aes(xintercept = 14 * 24), linetype = "dashed") + 
  scale_colour_manual(values = c("black", "red")) +
  labs(y = "Optical Density (OD)", x = "Elapsed Time (h)", title = "Tubes") +
  facet_grid(cols = vars(as.factor(Tube))) +
  theme_bw()

 TargetData %>%
  filter(grepl("od-", key)) %>%
  ggplot(data = .) +
  geom_point(aes(x = time, y = value, colour = as.factor(str_detect(key, "680"))), size = 0.5) +
   coord_cartesian(ylim = c(-0.1,0.1)) +
  geom_vline(aes(xintercept = 7 * 24), linetype = "dashed") + 
  geom_vline(aes(xintercept = 14 * 24), linetype = "dashed") + 
  scale_colour_manual(values = c("black", "red")) +
  labs(y = "Optical Density (OD)", x = "Elapsed Time (h)", title = "Tubes") +
  facet_grid(cols = vars(as.factor(Tube))) +
  theme_bw() 


```

##### Generate par_ue column with rows aligned with OD measures
>Pivot_wider to get actinic-lights data aligned with relevant sensor data. Need to include arrange(Filename, time, Tube) to keep things aligned! Need to group_by and/or reorder rows appropriately; Be Careful

```{r pivot_wider MCMIX004}
#possible issue with data structure; there are multiple values for some of the rows of actinic light columns, so the column becomes a list.
#Can add  values_fn = 
#to arbitrarily take the max or min etc. element of the list; but there might be a wider problem here when importing multiple files

TargetDataWide <- TargetData %>%
  pivot_wider(names_from = key, values_from = value, values_fn = list(value = max)) %>%
  arrange(Filename, MC, Tube, time)

TargetDataWide[1:10,]

rm(TargetData)
```

>Actinic light values do not align time wise with OD measures.

>Interpolate NA in actinic light columns from last observation, arrange by MC & Tube Then generate Actinic_par summary column. If multiple lights are activated, this chunk will give the summed par of all different colours for the tube. If a single actinic light is activated per tube, this gives the par for that tube. Filter rows where !is.na(Actinic_par) to check for incorrect row sums.

> Interpolation for Sine is not necessarily appropriate interpolation for Square photoregime; issues with propagating last Actinic_par of afternoon through evening, or back-casting first Actinic_par of morning.

>Small glitching adding actinic_light values for tubes where
actinic_light column should be 0; issue with interpolation we think.

```{r interpolate and summarize actinic_par by tube MCMIX004}
#http://publish.illinois.edu/spencer-guerrero/2014/12/11/2-dealing-with-missing-data-in-r-omit-approx-or-spline-part-1/
#https://dplyr.tidyverse.org/dev/articles/colwise.html
#Interpolation causes problems with final rows that repeat last value.

interpolate <- function(x){zoo::na.locf(x, na.rm = FALSE, fromLast = FALSE, type = "l", maxgap = Inf)}

#possible problem with actinic_par for MC data b/c actinic-lights.light1 = NA generates actinic_par of 0 b/c in rowSums na.rm = TRUE, which treats NA as 0.
#possibly not a big problem but watch for bugs
#na.rm = FALSE fails to run
TargetDataWide <- TargetDataWide %>%
  group_by(Tube) %>%
  arrange(Filename, MC, Tube, time) %>%
  mutate(across(.cols = starts_with("actinic-lights.light"), .fns = interpolate)) %>%
  ungroup() %>%
  mutate(Actinic_par = rowSums(.[grep("actinic-lights.light", names(.))], na.rm = TRUE)) %>%
  filter(!is.na(Actinic_par)) %>%
   dplyr::select(!contains("actinic-lights.light"))


TargetDataWide[1:10,]
```

>Now that Actinic_par is aligned with each row, coalesce
od-sensors-X.od-720 and od-sensors-X.od-680 into 2 columns, b/c 'Tube'is already a column, so no need to identify tube X in
od-sensors-X.od-680 columns. This might cause problems later matching OD measures to actinic light colours.

```{r consolidate OD MCMIX004}
TargetDataWide <- TargetDataWide  %>%
   mutate(OD680 = rowSums(.[grep("od-680", names(.))], na.rm = TRUE),
          OD720 = rowSums(.[grep("od-720", names(.))], na.rm = TRUE)) %>%
   dplyr::select(!contains("od-sensors"))

TargetDataWide[1:10,]

```

#### Merge Data with meta data
```{r metadata merge MCMIX004}
#This generates 'NA' values for ~1,000,000 rows of 'O2'; possibly the temperature rows?
TargetDataMeta <- left_join(x = TargetDataWide, y= CultureCatalog, by = c("MC", "Tube"))

TargetDataMeta

rm(TargetDataWide)
```

#### Prelim plot with culture meta (MCMIX004)
```{r second prelim plot MCMIX004}

  TargetDataMeta %>% ggplot() +
  geom_point(aes(x = time, y = OD680, colour = as.factor(WL)), size = 0.15) +
  geom_point(aes(x = time, y = OD720), size = 0.15, colour = "orange") +
  #geom_point(aes(x = time, y = Actinic_par/1000),  colour = "orange", size = 0.0001) +
  geom_vline(aes(xintercept = 7 * 24), linetype = "dashed") + 
  geom_vline(aes(xintercept = 14 * 24), linetype = "dashed") + 
  #scale_x_continuous(breaks=seq(0, 800, by = 125)) +
  #coord_cartesian(xlim = c(-10, 800)) +
 # scale_colour_manual(values = MCMIXColours) +
  #labs(y = "Optical Density (OD)", x = "Elapsed Time (h)", subtitle = "Growth Light (µE); Strain; SampleID; Tube") +
  facet_grid(rows = vars(as.factor(Plastic_Present)), cols = vars(as.factor(Tube), Nutrient_Present, Culture_Present)) +
  theme_bw()

```


### MCMIX006
#### Set Target File and read in csv
```{r set TargetFile and read csv MCMIX006}

TargetFile <- "20240927_Plastic_MCMIX006.csv"


# This opens a compressed file without decompressing.
TargetData <- read_csv(unz(DataIn, TargetFile),  skip = Skip, id = "Path", col_names = c("key", "time", "abs-time", "value")) %>% 
  separate(col =Path, into = c("Path", "Filename"), sep = ":") 



TargetFileName <- str_remove(string = TargetFile, pattern = ".csv")

TargetFileName

```

#### Filter superfluous rows, Add ToD column, extract ExpDate, extract MC,
```{r tidy MultiDataTarget MCMIX006}
#filter superfluous rows to simplify later pivot
TargetData <- TargetData %>%
  filter(str_detect(key, "od-720|od-680|actinic-lights.light"))
#head(TargetData)

# some files have "V5"  instead of "value" column
TargetData <- TargetData %>%
  dplyr::select(key, time, `abs-time`, value, Filename) %>% #, CDateTime
  mutate(Tube = str_extract(key, "-[:digit:]"),
         Tube = as.numeric(str_extract(Tube, "[:digit:]")),
         abs_time = mdy_hms(`abs-time`)) %>%
  dplyr::select(-`abs-time`)


#extract StartHour dynamically from first row of abs_time and display for cross check
StartHour <- as.numeric(format(TargetData$abs_time[1], format = "%H"))
StartHour
 
StartDate <- TargetData$abs_time[1]
StartDate

#Generate ToD as mod 24 of time + StartHour
TargetData <- TargetData %>%
  mutate(ToD = (time + StartHour) %% 24,
         Day = round((time/24), digits = 0))
#had trouble extracting 'Day' from abs_time, only got 3 unique values
  
#extract ExpDate for matching with Catalog
TargetData <- TargetData %>% 
    mutate(ExpDate = str_extract(Filename, "202[:digit:][:digit:][:digit:][:digit:][:digit:]_"),
           ExpDate = ymd(str_extract(ExpDate, "202[:digit:][:digit:][:digit:][:digit:][:digit:]")))

#extract MC for matching with Catalog
#fixed to run with MC or MCMIX.
#fixed to work for RUN #'s > 100. MC extracted from filename when Runs >99 returns the MC name and RUN #, this removes '_RUN#'
TargetData <- TargetData %>% 
    mutate(MC = str_extract(Filename, "MC.*[:digit:][:digit:][:digit:]"),
           MC = str_remove(MC, pattern = "\\_.*"))

```

#### Create preliminary plot for TargetData MCMIX006
##### Plots all OD values. True detection is OD680 and false detection is OD720.
```{r prelim plot MCMIX006}

  TargetData %>%
  filter(grepl("od-", key)) %>%
  ggplot(data = .) +
  geom_point(aes(x = time, y = value, colour = as.factor(str_detect(key, "680"))), size = 0.5) +
  geom_vline(aes(xintercept = 7 * 24), linetype = "dashed") + 
  geom_vline(aes(xintercept = 14 * 24), linetype = "dashed") + 
  scale_colour_manual(values = c("black", "red")) +
  labs(y = "Optical Density (OD)", x = "Elapsed Time (h)", title = "Tubes") +
  facet_grid(cols = vars(as.factor(Tube))) +
  theme_bw()

 TargetData %>%
  filter(grepl("od-", key)) %>%
  ggplot(data = .) +
  geom_point(aes(x = time, y = value, colour = as.factor(str_detect(key, "680"))), size = 0.5) +
   coord_cartesian(ylim = c(-0.1,0.1)) +
  geom_vline(aes(xintercept = 7 * 24), linetype = "dashed") + 
  geom_vline(aes(xintercept = 14 * 24), linetype = "dashed") + 
  scale_colour_manual(values = c("black", "red")) +
  labs(y = "Optical Density (OD)", x = "Elapsed Time (h)", title = "Tubes") +
  facet_grid(cols = vars(as.factor(Tube))) +
  theme_bw() 


```


##### Generate par_ue column with rows aligned with OD measures
```{r pivot_wider MCMIX006}
#possible issue with data structure; there are multiple values for some of the rows of actinic light columns, so the column becomes a list.
#Can add  values_fn = 
#to arbitrarily take the max or min etc. element of the list; but there might be a wider problem here when importing multiple files

TargetDataWide <- TargetData %>%
  pivot_wider(names_from = key, values_from = value, values_fn = list(value = max)) %>%
  arrange(Filename, MC, Tube, time)

TargetDataWide[1:10,]

rm(TargetData)
```

```{r interpolate and summarize actinic_par by tube MCMIX006}
#http://publish.illinois.edu/spencer-guerrero/2014/12/11/2-dealing-with-missing-data-in-r-omit-approx-or-spline-part-1/

#https://dplyr.tidyverse.org/dev/articles/colwise.html

#Interpolation causes problems with final rows that repeat last value.

interpolate <- function(x){zoo::na.locf(x, na.rm = FALSE, fromLast = FALSE, type = "l", maxgap = Inf)}

#possible problem with actinic_par for MC data b/c actinic-lights.light1 = NA generates actinic_par of 0 b/c in rowSums na.rm = TRUE, which treats NA as 0.
#possibly not a big problem but watch for bugs
#na.rm = FALSE fails to run
TargetDataWide <- TargetDataWide %>%
  group_by(Tube) %>%
  arrange(Filename, MC, Tube, time) %>%
  mutate(across(.cols = starts_with("actinic-lights.light"), .fns = interpolate)) %>%
  ungroup() %>%
  mutate(Actinic_par = rowSums(.[grep("actinic-lights.light", names(.))], na.rm = TRUE)) %>%
  filter(!is.na(Actinic_par)) %>%
   dplyr::select(!contains("actinic-lights.light"))


TargetDataWide[1:10,]
```

```{r consolidate OD MCMIX006}
TargetDataWide <- TargetDataWide  %>%
   mutate(OD680 = rowSums(.[grep("od-680", names(.))], na.rm = TRUE),
          OD720 = rowSums(.[grep("od-720", names(.))], na.rm = TRUE)) %>%
   dplyr::select(!contains("od-sensors"))

TargetDataWide[1:10,]

```

#### Merge Data with meta data

```{r metadata merge MCMIX006}
#This generates 'NA' values for ~1,000,000 rows of 'O2'; possibly the temperature rows?
TargetDataMeta <- left_join(x = TargetDataWide, y= CultureCatalog, by = c("MC", "Tube"))

TargetDataMeta

rm(TargetDataWide)
```
#### Prelim plot with culture meta data (MCMIX006)
```{r second prelim plot MCMIX006}

  TargetDataMeta %>% ggplot() +
  geom_point(aes(x = time, y = OD680, colour = as.factor(WL)), size = 0.15) +
  geom_point(aes(x = time, y = OD720), size = 0.15, colour = "orange") +
  #geom_point(aes(x = time, y = Actinic_par/1000),  colour = "orange", size = 0.0001) +
  geom_vline(aes(xintercept = 7 * 24), linetype = "dashed") + 
  geom_vline(aes(xintercept = 14 * 24), linetype = "dashed") + 
  #scale_x_continuous(breaks=seq(0, 800, by = 125)) +
  #coord_cartesian(xlim = c(-10, 800)) +
  #scale_colour_manual(values = MCMIXColours) +
  #labs(y = "Optical Density (OD)", x = "Elapsed Time (h)", subtitle = "Growth Light (µE); Strain; SampleID; Tube") +
  facet_grid(rows = vars(as.factor(Plastic_Present)), cols = vars(as.factor(Tube), Nutrient_Present, Culture_Present)) +
  theme_bw()

```






XXXLook at what you want for filtering
Fix meta catalog
do for both of the files
fix labels
as a formal practice segregate plastic (but fine for this project)
remove extra columns from meta





# Plastic Analyses
```{r plastic percent change}
# make a new data set to not mess up the original
plastic <- CultureCatalog


# Calculate percentage change
plastic$perc_change <- (plastic$Final_Plastic_Weight_mg - plastic$Initial_Plastic_Weight_mg) / plastic$Initial_Plastic_Weight_mg * 100

plastic
```
```{r plastic percent change graph}
ggplot(plastic, aes(x = as.factor(Culture_Present), y = perc_change, fill = as.factor(Nutrient_Present))) +
  geom_boxplot() +
  labs(title = "Plastic Degradation Across Conditions", x = "Culture Present", y = "Percentage Change (%)") +
  theme_minimal()
```


# Microbial Suspension and Wall Growth Analyses

```{r}
# Scatter plot: microbial growth vs plastic degradation
ggplot(plastic, aes(x = perc_change, y = Wall_Growth_RFU, color = as.factor(Nutrient_Present))) +
  geom_point() +
  labs(title = "Plastic Degradation vs Microbial Growth", x = "Percentage Plastic Change (%)", y = "Wall Growth RFU") +
  theme_minimal()
```

```{r}
# ANOVA model
anova_model <- aov(perc_change ~ Culture_Present * Nutrient_Present * Temperature_C, data = plastic)
summary(anova_model)

# plot of the interactions
interaction.plot(plastic$Temperature_C, plastic$Culture_Present, plastic$perc_change,
                 col = 1:2, xlab = "Temperature", ylab = "Plastic Degradation (%)",
                 main = "Interaction Between Temperature and Culture")

```

XXX
```{r}
# Correlation matrix
library(reshape2)
cor_data <- cor(plastic[, c("perc_change", "Wall_Growth_RFU", "Suspension_RFU")], use = "complete.obs")
cor_melt <- melt(cor_data)

# Heatmap
ggplot(cor_melt, aes(Var1, Var2, fill = value)) +
  geom_tile() +
  scale_fill_gradient2(low = "blue", high = "red", mid = "white", midpoint = 0) +
  labs(title = "Correlation Heatmap", fill = "Correlation") +
  theme_minimal()

```




## converting the variables to factor type
```{r making variables factors}
# Convert grouping variables to factors 
plastic$Plastic_Present <- as.factor(plastic$Plastic_Present)
plastic$Nutrient_Present <- as.factor(plastic$Nutrient_Present)
plastic$Culture_Present <- as.factor(plastic$Culture_Present)
```




# Extra analyses
```{r percent loss or gain plastic}
graph_data <- CultureCatalog %>%
  mutate(Weight_Loss_Percent = ((Initial_Plastic_Weight_mg - Final_Plastic_Weight_mg) / Initial_Plastic_Weight_mg) * 100)

graph_data$Culture_Present <- as.factor(graph_data$Culture_Present)

head(graph_data)



```

# Yvanna
```{r}
data_long<-CultureCatalog%>%
  pivot_longer(cols=c(Initial_Plastic_Weight_mg,Final_Plastic_Weight_mg),names_to = "Weight_Type",values_to = "Weight")
data_long

#putting them in the right order
data_long$Weight_Type <- factor(data_long$Weight_Type, levels = c("Initial_Plastic_Weight_mg", "Final_Plastic_Weight_mg"))
```
```{r}
ggplot(data_long,
       aes(x=Tube,y=Weight,fill=Weight_Type))+
  geom_bar(stat="identity",position="dodge")+
  scale_fill_manual(
    values=c("Initial_Plastic_Weight_mg"="lightblue",
             "Final_Plastic_Weight_mg"="red2"))+
  labs(x="Tube",y="Weight(mg)",fill="Weight Type")+
  facet_wrap(~ Temperature_C, labeller = labeller(Temperature_C = label_both)) +
  theme_minimal() +
  theme(
    strip.background = element_rect(fill = "lightgray", color = "black"), 
    strip.text = element_text(size = 9, face = "bold"), 
    panel.border = element_rect(color = "black", fill = NA, linewidth = 0.8)) 
  
```



```{r}


ggplot(graph_data, aes(x = Temperature_C, y = Wall_Growth_RFU, colour = Culture_Present)) +
  geom_point(size = 3, alpha = 0.7) +
  labs(x = "Temperature (°C)", y = "Wall Growth RFU", 
       title = "Scatter Plot of Wall Growth RFU vs Temperature") +
  theme_minimal()
ggplot(graph_data, aes(x = Temperature_C, y = Suspension_RFU, colour = Culture_Present)) +
  geom_point(size = 3, alpha = 0.7) +
  labs(x = "Temperature (°C)", y = "Suspension RFU", 
       title = "Scatter Plot of Suspension RFU vs Temperature") +
  theme_minimal()



```



# Andrew
I will be running an ANOVA (one way) on change in plastic mass data (initial - end) for experimental groups
(5 tubes each) in 15 and 25 degrees celsius.

This will tell us whether the mean '% change in plastic weight' is significantly different between the 15 and 25 degree groups


```{r}
filtered_dataplasticpresent<-CultureCatalog[CultureCatalog$Plastic_Present != 0,]
```


```{r}
onewayanova_mass<-aov((filtered_dataplasticpresent$Initial_Plastic_Weight_mg-filtered_dataplasticpresent$Final_Plastic_Weight_mg)~filtered_dataplasticpresent$Temperature_C)
summary(onewayanova_mass)
```

Factorial ANOVA using temperature, culture presence, and nutrient presence as independent variables.
This will tell us whether any one of these variables or some combination have a statistically significant effect on change in plastic mass.
```{r}
factorialanova_mass<-aov((filtered_dataplasticpresent$Initial_Plastic_Weight_mg-filtered_dataplasticpresent$Final_Plastic_Weight_mg)~filtered_dataplasticpresent$Temperature_C*filtered_dataplasticpresent$Culture_Present*filtered_dataplasticpresent$Nutrient_Present)
summary(factorialanova_mass)
```

One way ANOVAs for chlorophyll assay readings using temp as independent variable (one ANOVA for wall growth and another for suspension)...
```{r}
onewayanova_rfuwall<-aov(CultureCatalog$Wall_Growth_RFU~CultureCatalog$Temperature_C)
summary(onewayanova_rfuwall)
onewayanova_rfususpension<-aov(CultureCatalog$Suspension_RFU~CultureCatalog$Temperature_C)
summary(onewayanova_rfususpension)
```
Factorial ANOVAs for chlorophyll assay readings using temp, nutrient presence, plastic presence and culture presence as independent variables...
```{r}
factorialanova_rfuwall<-aov(CultureCatalog$Wall_Growth_RFU~CultureCatalog$Temperature_C*CultureCatalog$Culture_Present*CultureCatalog$Nutrient_Present*CultureCatalog$Plastic_Present)
summary(factorialanova_rfuwall)
factorialanova_rfususpension<-aov(CultureCatalog$Suspension_RFU~CultureCatalog$Temperature_C*CultureCatalog$Culture_Present*CultureCatalog$Nutrient_Present*CultureCatalog$Plastic_Present)
summary(factorialanova_rfususpension)
```










# more analyses 
XXX fix these
```{r}
# Summary statistics for Final_Plastic_Weight_mg
plastic %>%
  group_by(Culture_Present, Nutrient_Present, Plastic_Present, Temperature_C) %>%
  summarise(
    mean_wall_rfu = mean(Wall_Growth_RFU, na.rm = TRUE),
    mean_susp_rfu = mean(Suspension_RFU, na.rm = TRUE)
  )

```

```{r}
# Add a column for Percent Weight Loss
plastic <- plastic %>%
  mutate(
    Percent_Weight_Loss = ((Initial_Plastic_Weight_mg - Final_Plastic_Weight_mg) / Initial_Plastic_Weight_mg) * 100
  )
plastic <- plastic %>%
  mutate(
    Temperature_C = as.factor(Temperature_C),
    Culture_Present = as.factor(Culture_Present),
    Nutrient_Present = as.factor(Nutrient_Present)
  )


# Two-way ANOVA for Percent Weight Loss
weight_loss_anova <- aov(Percent_Weight_Loss ~ Temperature_C * Culture_Present * Nutrient_Present, data = plastic)
summary(weight_loss_anova)

# Tukey HSD for pairwise comparisons
TukeyHSD(weight_loss_anova)

```

```{r}
# Two-way ANOVA for Wall_Growth_RFU
rfu_anova <- aov(Wall_Growth_RFU ~ Temperature_C * Culture_Present * Nutrient_Present, data = plastic)
summary(rfu_anova)

# Linear regression for RFU vs Percent Weight Loss
rfu_regression <- lm(Percent_Weight_Loss ~ Wall_Growth_RFU + Suspension_RFU, data = plastic)
summary(rfu_regression)

# Scatterplot of RFU vs Percent Weight Loss
ggplot(plastic, aes(x = Wall_Growth_RFU, y = Percent_Weight_Loss, color = Temperature_C)) +
  geom_point() +
  geom_smooth(method = "lm", se = TRUE) +
  labs(x = "Wall Growth RFU", y = "Percent Weight Loss")

```

```{r}
# T-test for weight loss by temperature
t.test(Percent_Weight_Loss ~ Temperature_C, data = plastic)

# Boxplot for Percent Weight Loss by Temperature
ggplot(plastic, aes(x = factor(Temperature_C), y = Percent_Weight_Loss)) +
  geom_boxplot() +
  labs(x = "Temperature (°C)", y = "Percent Weight Loss")

```




```{r}
# Calculate Summary Statistics
summary_stats <- plastic %>%
  group_by(Temperature_C, Culture_Present, Nutrient_Present) %>%
  summarise(
    mean_loss = mean(Percent_Weight_Loss, na.rm = TRUE),
    se_loss = sd(Percent_Weight_Loss, na.rm = TRUE) / sqrt(n()),
    .groups = "drop")

# Bar Plot of Average Percent Weight Loss
ggplot(summary_stats, aes(x = factor(Temperature_C), y = mean_loss, fill = Culture_Present)) +
  geom_bar(stat = "identity", position = position_dodge(0.9)) +
  geom_errorbar(aes(ymin = mean_loss - se_loss, ymax = mean_loss + se_loss),
                position = position_dodge(0.9), width = 0.2) +
  labs(x = "Temperature (°C)", y = "Mean Percent Weight Loss", fill = "Culture Present") +
  theme_minimal()

# mean loss having negative means that it actually grew

# Interaction Plot
interaction.plot(plastic$Temperature_C, plastic$Nutrient_Present, plastic$Percent_Weight_Loss)

```

```{r}
ggplot(summary_stats, aes(x = factor(Temperature_C), y = mean_loss, fill = Culture_Present)) +
  geom_bar(stat = "identity", position = position_dodge(0.9)) +
  geom_errorbar(aes(ymin = mean_loss - se_loss, ymax = mean_loss + se_loss),
                position = position_dodge(0.9), width = 0.2) +
  labs(x = "Temperature (°C)", y = "Mean Percent Weight Loss", fill = "Culture Present") +
  facet_wrap(~ Nutrient_Present) +
  theme_minimal()

```

```{r}
# Bar Plot with Facet Labels and Borders
ggplot(summary_stats, aes(x = factor(Temperature_C), y = mean_loss, fill = Culture_Present)) +
  geom_bar(stat = "identity", position = position_dodge(0.9)) +
  geom_errorbar(aes(ymin = mean_loss - se_loss, ymax = mean_loss + se_loss),
                position = position_dodge(0.9), width = 0.2) +
  labs(
    x = "Temperature (°C)",
    y = "Mean Percent Weight Loss",
    fill = "Culture Present"
  ) +
  facet_wrap(~ Nutrient_Present, labeller = labeller(Nutrient_Present = label_both)) +
  theme_minimal() +
  theme(
    strip.background = element_rect(fill = "lightgray", color = "black"), 
    strip.text = element_text(size = 12, face = "bold"), # Style facet labels
    panel.border = element_rect(color = "black", fill = NA, linewidth = 0.8) 
  )


```




### looking at the negative loss (gain) category
```{r negative loss (gain)}
# Filter the dataset for the negative mean category
negative_mean_data <- plastic %>%
  filter(Percent_Weight_Loss < 0)
negative_mean_data
```

# Results


## Pictures 
XXX make sure that yvanna cites herself

# Discussion




XXX need to do every single read me in every single folder


# References


